Staged Abstract Interpreters

### 2019-02-27 paper/10_intro.tex
update

\begin{lstlisting}
def power(b: Rep[Int], e: Int): Rep[Int] = 
  if (e == 0) 1 else e * power(b, e-1)
\end{lstlisting}

@@

\begin{lstlisting}
def power5(b: Int) = b * b * b * b * b
\end{lstlisting}

@@

Now we have seen two orthogonal transformations of interpreters: by
specialization, we can refactor interpreters to code generators; by approximation,
we can refactor interpreters to static analyzers. As an intellectual quest, it
is natural to ask --- starting from an interpreter, \textbf{can we derive a
sound abstract interpreter but it works like a code generator which produces fast
analysis code?} Conceptually, it is similar to the concrete setting: for an
abstract interpreter $\widehat{\texttt{eval}}: \texttt{Expr} \to
\widehat{\texttt{Env}} \to \widehat{\texttt{Value}}$, which takes a program, an abstract
environment and returns abstract values, we would like to specialize it with
respect to a program $e_0$ and produce a compiled analysis
$\widehat{\texttt{eval}}_{\texttt{e0}} : \widehat{\texttt{Env}} \to
\widehat{\texttt{Value}}$, such that $
\widehat{\texttt{eval}}_{\texttt{e0}}(\widehat{\rho}) =
\widehat{\texttt{eval}}(e_0)(\widehat{\rho})$, but the compiled one leaves the
abstract environment $\widehat{\rho}$ as remaining input.

@@

Besides the intellectual merit shown in the confluence diagram, this paper
contributes to a practical problem in static analysis: \textbf{can we speed up a
static analyzer without any intrusive changes and soundness compromises?} Static
analysis is known as a trade-off game between \textit{precision} and
\textit{efficiency}, we argue that applying meta-programming (multi-stage
programming, in particular) with abstract interpreters is an effective approach
to improve the latter one while the former one is untouched. Specialization
removes the overhead of abstractions introduced by high-level programming, but
not changes the meaning of the program -- in this case, the abstract semantics
implemented by the analyzer. Indeed, specializing analysis with respect to a
program by partial evaluation \cite{damian1999partial, amtoft1999partial,
Boucher:1996:ACN:647473.727587, ashley:practical} is not a new idea, despite the
fact that most of the previous work are ad-hoc on a particular analysis or
requiring significant changes to the analysis. For example, abstract compilation
\cite{Boucher:1996:ACN:647473.727587} requires the whole analyzer to be
rewritten as closure generation from. In this paper, we show a systematic
approach to meta-programming achieves the performance goal with less intrusive
changes; in fact, a case study (Section~\ref{cs_ac}) comparing with abstract
compilation shows, by utilizing type-based stage annotations, the analyzer
program does not need any changes.

>>>>

\begin{lstlisting}
def power(b: Rep[Int], e: Int): Rep[Int] = 
  if (e == 0) 1 else b * power(b, e - 1)
\end{lstlisting}

@@

\begin{lstlisting}
def power5(b: Int): Int = b * b * b * b * b
\end{lstlisting}

@@

Now we have seen two orthogonal transformations of interpreters: by
specialization, we can refactor interpreters to code generators; by approximation,
we can refactor interpreters to static analyzers. As an intellectual quest, it
is natural to ask --- starting from an interpreter, \textit{can we derive a
sound abstract interpreter while it produces fast code that does the analysis?}
Conceptually, it is similar to the concrete setting: for an
abstract interpreter $\widehat{\texttt{eval}}: \texttt{Expr} \to
\widehat{\texttt{Env}} \to \widehat{\texttt{Value}}$, which takes a program, an abstract
environment and returns abstract values, we would like to specialize it with
respect to a program $e_0$ and produce a compiled analysis
$\widehat{\texttt{eval}}_{\texttt{e0}} : \widehat{\texttt{Env}} \to
\widehat{\texttt{Value}}$, such that $
\widehat{\texttt{eval}}_{\texttt{e0}}(\widehat{\rho}) =
\widehat{\texttt{eval}}(e_0)(\widehat{\rho})$, but the compiled one leaves the
abstract environment $\widehat{\rho}$ as remaining input.

@@

Besides the intellectual merit shown in the confluence diagram, this paper
contributes to a practical problem in static analysis: \textit{can we speed up a
static analyzer without any intrusive changes and soundness compromises?} Static
analysis is known as a trade-off game between \textit{precision} and
\textit{efficiency}, we argue that applying meta-programming (multi-stage
programming, in particular) with abstract interpreters is an effective approach
to improve the latter one while the former one is untouched. Specialization
removes the overhead of abstractions introduced by high-level programming, but
not changes the meaning of the program -- in this case, the abstract semantics
implemented by the analyzer. Indeed, specializing analysis with respect to a
program by partial evaluation \cite{damian1999partial, amtoft1999partial,
Boucher:1996:ACN:647473.727587, ashley:practical} is not a new idea, despite the
fact that most of the previous work are ad-hoc on a particular analysis or
requiring significant changes to the analysis. For example, abstract compilation
\cite{Boucher:1996:ACN:647473.727587} requires the whole analyzer to be
rewritten as closure generation from. In this paper, we show a systematic
approach to meta-programming achieves the performance goal with less intrusive
changes; in fact, a case study (Section~\ref{cs_ac}) comparing with abstract
compilation shows, by utilizing type-based stage annotations, the analyzer
program does not need any changes.

### 2019-02-27 paper/21_bg_interp.tex
update

We will give the concrete semantics using a big-step definitional
interpreter. The interpreter is a recursive function that takes the program AST,
environment and store, and returns the evaluated value and the accompanied
store. The environment is a mapping from identifiers to addresses, and the store
is a mapping from addresses to values. We use the store to model recursion and
mutation in concrete semantics; it is also useful for polyvariant analysis. This
environment-and-store-passing style big-step interpreter is standard and can
also be obtained by refunctionalizing \cite{DBLP:conf/ppdp/AgerBDM03,
Wei:2018:RAA:3243631.3236800} a small-step CESK machine
\cite{DBLP:conf/popl/FelleisenF87}.

@@

>>>>

The abstract syntax we present in fact can be seen as a deep embedding of the
language -- we use datatypes to represent programs. This design choice makes us
easily add different interpretations over the AST; with the inheritance and
overriding mechanism in Scala, we may also add new language construct and reuse
the existing interpretation.

\iffalse
We will give the concrete semantics using a big-step definitional
interpreter. The interpreter is a recursive function that takes the program AST,
environment and store, and returns the evaluated value and the accompanied
store. The environment is a mapping from identifiers to addresses, and the store
is a mapping from addresses to values. We use the store to model recursion and
mutation in concrete semantics; it is also useful for polyvariant analysis. This
environment-and-store-passing style big-step interpreter is standard and can
also be obtained by refunctionalizing \cite{DBLP:conf/ppdp/AgerBDM03,
Wei:2018:RAA:3243631.3236800} a small-step CESK machine
\cite{DBLP:conf/popl/FelleisenF87}.
\fi

\subsection{Monads in Scala}

In Scala, thanks for its nature that combines functional programming and
object-oriented programming, we may use traits and objects to implement monads.
A trait poses monadic operations as abstract methods that objects extending it
should implement. An object encapsulates the data and implements monadic methods
that operate on these data. The monadic @bind@ operation is usually called
@flatMap@ in Scala. For example, the following trait @Monad@ declares the
@return@ and @flatMap@ operation:

\begin{lstlisting}
  trait Monad[F[_]] {
    def return[A](a: => A): F[A]
    def flatMap[A,B](fa: F[A])(f: A => F[B]): F[B]
  }
\end{lstlisting}

An object may encapsulate the data of type @F[A]@ and only provide
@flatMap(f: A => F[B]): F[B]@. This allows us to utilize the for-comprehensions in
Scala, which provide special syntactic support for moandic @flatMap@.
For example, a list object of element type @A@ is a monad and it implements the
methods @flatMap[B](f: A => List[B]): List[B]@ and @map[B](f: A => B): List[B]@.
To compute the Cartesian product of two lists of numbers, we may directly call
@flatMap@ and pass an anonymous function that internally calls the monadic
@return@ operation (the @List@ constructor in this case):

\begin{lstlisting}
  val xs = List(1, 2); val ys = List(4, 5)
  xs.flatMap(x => ys.flatMap(y => List((x, y)))) // List((1,4), (1,5), (2,4), (2,5))
\end{lstlisting}

Or, we can use for-comprehensions that computes the same value:

\begin{lstlisting}
  for {x <- xs; y <- ys} yield (x, y) // List((1,4), (1,5), (2,4), (2,5))
\end{lstlisting}

The Scala compiler will translate the above for-comprehension expression into
an equivalent one using @flatMap@ and @map@ ~\cite{scala_spec}:

\begin{lstlisting}
  xs.flatMap { case x => ys.map { case y => (x, y) } }
\end{lstlisting}

In the rest of the paper, we will use for-comprehension syntax for monadic
operations, the implementation of monads and monad transformers essentially
borrows the ground-truth from Haskell.

@@

Monads provide an elegant and extensible way to build
interpreters\cite{DBLP:conf/popl/LiangHJ95, Steele:1994:BIC:174675.178068,
Wadler:1992:EFP:143165.143169}, and recently the approach has been extended to
abstract interpreters \cite{Sergey:2013:MAI:2491956.2491979,
DBLP:journals/pacmpl/DaraisLNH17}. In section, with the multi-stage programming
and abstract interpretation in mind, we use moandic operations to construct an
generic interface in the style of big-step definitional interpreter.
